---
title: "LINEAR REGRESSION IN R"
author: "Becky Njuguna"
date: "2024-01-22"
output: html_document
---
# Determining the predictors of 6 minute walking distance using linear regression.
Hard small. That is how a Kenyan would say something is "Ngumu kidogo", rounded off to the nearest English. 

This, for me, was not a little hard. It was " a lottle" hard. Learning a new language in your 30's is what Kenyan society would call "Masomo ya gumbaru" Regardless,or should I say "irregardless", I persevered. 

In 2023, I decided to start my data analytics journey. I chose to start by learning R as it is free, there are a lot of resources online and the R community is also very large. 
In the project below, I used a data set provided while doing an online R course on Coursera, "Linear Regression in R for Public Health" by Imperial College London. 



## Data Summary

```{r}
#Loading data set into R Studio
COPD <- read.csv("COPD_student_dataset.csv")
head (COPD)
```


The data set above has 101 entries. 

*X- Ordered entry numbers  
*ID- unique identifiers  
*Age- age in years  
*PackHistory-smoking history in pack years  
*COPDSEVERITY- Severity of COPD MILD, -1, MODERATE-2, SEVERE-3, VERY SEVERE -4    
*MWT1-First measure of distance walked in 6 minutes in metres  
*MWT2- Second measure of Distance walked in 6 minutes in metres  
*MWT1Best- Best walking distance in 6 minutes in metres  
*FEV1- Forced expiratory volume in one minute in metres  
*FEV1PRED- Predicted forced expiratoty volume in 1 minute in litres  
*FVC-Forced Vital Capacity in litres  
*FVCPRED - Predicted Forced Vital Capacity in litres  
*CAT- COPD Assesement Score  
*HAD- Hospital Anxiety and depression Score  
*SGRQ- St. George's Respiratory Questionnaire score.  
*AGEquartiles-Age Quartiles, 1-4  
*copd severity, 1-4
*smoking:no-0, yes-1  
*Diabetes:no-0, yes-1  
*muscular: no-0, yes-1  
*hypertension :no-0, yes-1  
*AtrialFib : no-0, yes-1  
*IHD: no-0, yes-1  


## Good Practise  
## Next, we need to inspect for missing and outlier values. Fist, I downloaded the Hmsic that allows me to do this.   
 
```{r}
#Installing the Hmsic package
install.packages("Hmisc", repos = "http://cran.r-project.org")
library(Hmisc)
#Description of individual variables
describe(COPD)
```

# This shows that only the response variabe has missing data. Furthermore, only 1 entry is missing.
# R's default is to exclude the missing data. 

# Next, I checked for the distributions of the continuous variables. 
```{r}
#Checking if any of the continuous predictor variables have narrow distributions
#Histogram to view the distribution of the variable, walking distance
hist(COPD$MWT1Best, main="Histogram of Distance walked in 6 minutes (MWT1Best)", xlab="MWT1Best", breaks=12) 
#Histogram to view distribution of variable, FEV1
hist(COPD$FEV1, main="Histogram of lung function (FEV1)", xlab="FEV1") 
#Histrogram to view the distribution of the variable FEV1PRED
hist(COPD$FEV1PRED, main="Histogram of lung function (FEV1PRED)", xlab="FEV1PRED") 
#Histrogram to view the distribution of the variable FVC
hist(COPD$FVC, main="Histogram of lung function (FVC)", xlab="FVC")
#Histrogram to view the distribution of the variable FVCPRED
hist(COPD$FVCPRED, main="Histogram of lung function (FVCPRED)", xlab="FVCPRED")
#Histrogram to view the distribution of the variable age
hist(COPD$AGE, main="Histogram of AGE", xlab="AGE")
#Histrogram to view the distribution of the variable packhistory
hist(COPD$PackHistory, main="Histogram of PackHistory", xlab="PackHistory")
```

#Examining the categorical variable to completeness

#Fist, I will create a new variable indicating the presence of at least one comorbidity (1) or complete absence (0)
```{r}
#First, all variables must be saved as factors
COPD$Diabetes <- as.factor(COPD$Diabetes)
class(COPD$Diabetes)
COPD$muscular <- as.factor(COPD$muscular)
class(COPD$muscular)
COPD$hypertension <- as.factor(COPD$hypertension)
class(COPD$hypertension)
COPD$AtrialFib <- as.factor(COPD$AtrialFib)
class(COPD$AtrialFib)
COPD$IHD <- as.factor(COPD$IHD)
class(COPD$IHD)
```

#Creating a new variable, comorbids
```{r}
#First make sure is same length as all other variables
comorbids <- length(COPD$X)
#Making it so that if with any comorbids, 1, if without, 0. 
comorbids [COPD$Diabetes==1| COPD$muscular==1| COPD$hypertension==1| COPD$AtrialFib==1| COPD$IHD== 1] <- 1
comorbids [is.na(comorbids)] <-0
comorbids <- factor(comorbids)
#Display Comorbids
print(comorbids)
#Summary of comorbids
summary(comorbids)
#Adding new comorbids variable into the data sets
COPD$comorbids <- comorbids
table(COPD$comorbids)
```

#This shows that there are 46 people with no comorbidities and 55 with at least one comorbidity. 







##Examine the relationship between your candidate continuous predictor variables**
```{r}
#Creating a correlation matrix for continuous variables. 
mylrmatrix<- COPD[, c("MWT1Best", "AGE" , "PackHistory", "FEV1", "FEV1PRED", "FVC", "FVCPRED")]
#Use pairwise complete observations, ignoring rows with missing values for the calculation
cor_matrix <- cor(mylrmatrix, use = "pairwise.complete.obs")
cor_matrix
#Correlation matrix to 2 decimal places
round(cor_matrix, 2)
#Correlation matrix plots
pairs(~AGE+PackHistory+ FEV1+ FEV1PRED+ FVC+ FVCPRED, data= COPD )
```

#Above shows that the lung functions are highly correlated with one another. 
#Therefore, FVCPRED is the best one to use given that the results of the pairwise corelation matrix are >-0.7 but < 0.7 

#Install the gmodels package (if not installed)
install.packages("gmodels")
# Load the gmodels package
library(gmodels)
#Crosstable for cartegorical variable
CrossTable (COPD$comorbids)


#Next, I ran linear regression model for each predictor against the dependent variable
```{r}
#Running a linear regression model for age and distance walked. 
MWT1Best_AGE <- lm(MWT1Best~AGE, data = COPD)
#Printing the model objects
lm(formula = MWT1Best~AGE, data = COPD )
#Summary of model
summary(MWT1Best_AGE)
#To view 95% confidence interval
confint(MWT1Best_AGE)
#Checking model assumptions
predictedvals <- predict(MWT1Best_AGE)
residualvals <- residuals(MWT1Best_AGE)
par(mfrow=c(2,2))
hist(residualvals, main = "Histogram of residuals", xlab = "Residuals")

```

```{r}
#Running a linear regression model for pack history and distance walked. 
MWT1Best_PackHistory <- lm(MWT1Best~PackHistory, data = COPD)
#Printing the model objects
lm(formula = MWT1Best~PackHistory, data = COPD )
#Summary of model
summary(MWT1Best_PackHistory)
#To view 95% confidence interval
confint(MWT1Best_PackHistory)
#Checking model assumptions
predictedvals <- predict(MWT1Best_PackHistory)
residualvals <- residuals(MWT1Best_PackHistory)
par(mfrow=c(2,2))
hist(residualvals, main = "Histogram of residuals", xlab = "Residuals")
```



```{r}
#Running a linear regression model for FVCPRED and distance walked. 
MWT1Best_FVCPRED <- lm(MWT1Best~FVCPRED, data = COPD)
#Printing the model objects
lm(formula = MWT1Best~FVCPRED, data = COPD )
#Summary of model
summary(MWT1Best_FVCPRED)
#To view 95% confidence interval
confint(MWT1Best_FVCPRED)
#Checking model assumptions
predictedvals <- predict(MWT1Best_FVCPRED)
residualvals <- residuals(MWT1Best_FVCPRED)
par(mfrow=c(2,2))
hist(residualvals, main = "Histogram of residuals", xlab = "Residuals")
```



```{r}
#Running a linear regression model for comorbids and distance walked. 
MWT1Best_comorbids <- lm(MWT1Best~comorbids, data = COPD)
#Printing the model objects
lm(formula = MWT1Best~comorbids, data = COPD )
#Summary of model
summary(MWT1Best_comorbids)
#To view 95% confidence interval
confint(MWT1Best_comorbids)
#Checking model assumptions
predictedvals <- predict(MWT1Best_comorbids)
residualvals <- residuals(MWT1Best_comorbids)
par(mfrow=c(2,2))
hist(residualvals, main = "Histogram of residuals", xlab = "Residuals")
```

```{r}
#Running a linear regression model for comorbids and distance walked. 
MWT1Best_comorbids <- lm(MWT1Best~comorbids, data = COPD)
#Printing the model objects
lm(formula = MWT1Best~comorbids, data = COPD )
#Summary of model
summary(MWT1Best_comorbids)
#To view 95% confidence interval
confint(MWT1Best_comorbids)
#Checking model assumptions
predictedvals <- predict(MWT1Best_comorbids)
residualvals <- residuals(MWT1Best_comorbids)
par(mfrow=c(2,2))
hist(residualvals, main = "Histogram of residuals", xlab = "Residuals")
```


** 3. Fit a simple linear regression model between outcome and each predictor **
```{r}
#LM between walking distance and gender
lr1AGE <- lm(MWT1Best~AGE, data= COPD)
summary(lr1AGE)
confint(lr1AGE)
#LM between walking distance and FEV1
lr2FEV1 <- lm(MWT1Best~FEV1, data= COPD)
summary(lr2FEV1)
confint(lr2FEV1)  
#LM between walking distance and FVC
lr3FVC <- lm(MWT1Best~FVC, data= COPD)
summary(lr3FVC)
confint(lr3FVC)    
#LM between walking distance and FEV1PRED
lr4FEV1PRED <- lm(MWT1Best~FEV1PRED, data= COPD)
summary(lr4FEV1PRED)
confint(lr4FEV1PRED)    
#LM between walking distance and FVCPRED
lr5FVCPRED <- lm(MWT1Best~FVCPRED, data= COPD)
summary(lr5FVCPRED)
confint(lr5FVCPRED) 
#LM between walking distance and presence of comorbidities
lr6comorbids <- lm(MWT1Best~comorbids, data = COPD)
summary(lr6comorbids)
confint(lr6comorbids) 
#LM between walking distance and packhistory
lr7PackHistory <- lm(MWT1Best~PackHistory, data = COPD)
summary(lr7PackHistory)
confint(lr7PackHistory) 

```

#Therefore pick FEV1 in the model over other lung function scored because the largest adjusted r in fev1 models explains fev to be the cause of most of the variance

```{r}
#Fitting the linear regression model for walking distance
multiplelmMWT1Best <- lm(MWT1Best~FEV1+ AGE + PackHistory+ factor(comorbids), data= COPD)
summary(multiplelmMWT1Best)
confint(multiplelmMWT1Best)
```

#Lack of change in the coefficients between the simple model and the multiple model should reassure you that collinearity is unlikely to be a problem in the model. 
#To examine the Variance inflation factors. 
```{r}
#Install car package for vif calculation
install.packages("car", repos = "http://cran.r-project.org")

library(car)
#Calculatimg the VIFs for the model
vif_values <- vif(multiplelmMWT1Best)
vif_values
```

#None of the VIFs is greater than 5 and therefore collinearity is not of concern.
```{r}

```
